The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-04-29 paper/ms.tex
number of scatterings

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{maxima.eps}
\end{center}
\caption{Position of the line maxima as a function of rotational
  velocity $V_{\rm max}$. Continuous (dashed) lines correspond to
  central (homogeneous) source distributions. A value of $x_{\rm
    max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumsvsvelocity}}  
\end{figure*}

We quantify the transition to a single peak by measuring the peak
maxima position as a function of the viewing angle and the rotational 
velocity. The results are presented in Figure
\ref{fig:maximumsvsvelocity}. There are three interesting features
that deserved attention. First For $|\mu|\sim 1.0$ and central sources
the two peaks have the same position regardless of the rotational
velocity. Second, in the same homogeneous models we see that for increasing
rotational velocities the line becomes single peaked, $x_{m}\sim
0$\kms as $|\mu|\sim 0.0$.

One possible explanation for the emergence of the single peak in the
homogeneous systems is that some photons close to the surface
(a sort of photosphere) can escape with a low number of scatterings,
allowing them to stay close to the line's center. Increasing the
rotational velocity $V_{\rm max}$ reduces the optical depth making the
photosphere region effectively larger, increasing the number of
photons escaping close to the line's center.

However, for the central emission the transition to a single peak is
almost never completed in the range of explored parameters. The absence
of a single peak phase could be partially explained by the absence of a
photosphere. In this case the average number of scatterings to escape
should remain close to constant as the rotational velocity
increases. The rise in intensity at the line's center could instead
mean that the scattering in a medium with bulk motion are inefficient
in driving photons outside that frequency region.

In order to explore our interpretation for these two scenarios we now 
quantify the effect of rotation on the number of scatterings.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f6.eps}
  \includegraphics[width=0.40\textwidth]{f7.eps}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel, escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel, escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
    slab is shown as a continuous    line. Different lines correspond
    to different rotational velocities for the homogeneous models.     
    \label{fig:efvsv}} 
\end{figure*}

In the case of dusty clouds, all changes in the average number
of scatterings presented in the previous subsection have a direct
bearing on the escape fraction. From the results we have just obtained
we can predict that for central source distributions the escape
fraction will barely change, while larger effects should be expected
for homogeneously distributed sources.

Left panel of Figure \ref{fig:efvsv} confirm these expectations. It shows the escape
fraction as a function of the rotational velocity. The curves for the
central source distribution stay flat, while for the homogeneous case
there is a clear rise with rotational velocity.  Rotation has a higher relative impact in
the models with low optical depth. For instance, in models with
$\tau_{\rm H}=10^5$, the static escape fraction is $0.26$ and increases
to $0.37$ for $V_{\rm max}=300$ \kms. Table \ref{table:escape} lists
all the values for the escape fraction.

In the right panel of Figure \ref{fig:efvsv} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{A}$ where $\tau_{A} = (1 - A)\tau_{a}$ , an approximation 
that is valid only in the limit $a\tau_{\rm   H}\gg 1$. The dashed lines in
 the right panel of Figure \ref{fig:efvsv} show the results for the different rotational
velocities for the homogeneous models. First of all we note that the escape fraction does not
increase significantly from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, This
 counter-intuitive result is a consequence of the transition into the 'extremely' opaque regime,
  which occurs roughly when $a\tau_{\rm{H}} > 10^{3}$
  e.g. \citep{Neufeld90}.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=0.40\textwidth]{nscattvsmu.eps}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  $\mu$. This is the general behavior for all models in particular this model
  is for $\tau=10^{5}$ and $V_{max}=300 km/s^{-1}$. \label{fig:Nscattvsmu}}    
\end{figure}

\begin{figure}
\begin{center}
    \includegraphics[width=0.40\textwidth]{f4.eps}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  rotational velocity. Continuous (dashed) lines represent an
  central (homogeneous) distribution of sources. \label{fig:Nscatt}}    
\end{figure}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{f5_1.eps}
  \includegraphics[width=0.45\textwidth]{f5_2.eps}
  \includegraphics[width=0.45\textwidth]{f5_3.eps}
  \includegraphics[width=0.45\textwidth]{f5_4.eps}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels correspond to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

The number of scatterings affects the escape frequency of a \ly photon.
In static cases, a large value of the optical depth correlates with a
high number of scatterings, increasing the probability of finding a
\ly photon far from the line's center. As a result the peak maxima
shifts from the center as the amount of neutral hydrogen
increases. This can be precisely quantified in the static slab
with central sources where the maxima position are related to
the optical depth and the temperature as $x_{m}=\pm 1.066(a\tau_{\rm
  H})^{1/3}$ \citep{Harrington73}. In the same model the average
number of scatterings depends only on the optical depth $\langle
N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}, in the case of homogeneously distributed
sources $\langle N_{\rm   scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.

In Figure \ref{fig:Nscatt} we can see the decrease of $\langle N_{\rm
  scatt}\rangle$ as $V_{\rm max}$ increases in the homogeneous
source distribution. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case.

For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this
represents factors of $(0.66, 0.59, 0.55)$ lower than the centrally
emitted photons. While for the homogeneous rotating sphere at 
$V_{\rm{max}}=300\kms$ we find that $\langle N_{\rm
  scatt}\rangle=(0.38, 0.34, 0.42)\tau_{\rm   H}$ which represents 
  factors of $(0.26, 0.34, 0.46)$ lower than the corresponding central
  case.

In order to gain a deeper understanding of these results and explain
the results for the maxima in the line morphology we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
such histograms in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.

>>>>

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{maxima.eps}
\end{center}
\caption{Position of the line maxima as a function of maximum
  rotational   velocity $V_{\rm max}$. Continuous (dashed) lines
  correspond to   central (homogeneous) source distributions. A value
  of $x_{\rm     max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumsvsvelocity}}
\end{figure*}

We measure the peak maxima poisition to quantify the transition from
double into single peak. We do this as a function of the viewing angle
$\theta$ and the maximum rotational velocity. The results are
summarized in Figure  \ref{fig:maximumsvsvelocity}. There are two
interesting features that deserve attention. First, for a viewing angle
parallel to the rotationa axis ($\mu\sim 1.0$) the maxima of all
models are similar regardless of the rotational velocity. Second, at a
viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) there
is a good fraction of models that become single peaked, a feature that
appears more frequently for homogeneously distributed sources when all
the other parameters are equal.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f6.eps}
  \includegraphics[width=0.40\textwidth]{f7.eps}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel, escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel, escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
    slab is shown as a continuous    line. Different lines correspond
    to different rotational velocities for the homogeneous models.     
    \label{fig:escape_fraction}} 
\end{figure*}

We now estimate the escape fraction $f_{\rm esc}$ for the dusty models. Our main
result is that we do not find any dependence with the viewing angle
but there is a strong dependence with the rotational velocity for the
homogeneous models.

The left panel in Figure \ref{fig:escape_fraction} summarizes our
results. The left panel shows the escape fraction as a function of the
rotational velocity. To construct this Figure we take into account all
the photons regardless of its outgoing direction. We observe that the
curves for the central source distribution stay flat, while for the
homogeneous case there is a clear rise with rotational velocity.
Rotation has a higher relative impact in the models with low optical
depth. For instance, in models with $\tau_{\rm H}=10^5$, the static
escape fraction is $0.26$ and increases to $0.37$ for $V_{\rm
  max}=300$ \kms. Table \ref{table:escape} lists all the values for
the escape fraction.

In the right panel of Figure \ref{fig:escape_fraction} we put these
results in the context of the analytic solution for the infinite slab
\citep{Neufeld90}. In Neufeld's set-up the analytic solution depends
uniquely on the product $(a\tau_{\rm   H})^{1/3}\tau_{A}$ where
$\tau_{A} = (1 - A)\tau_{a}$, valid only in the limit $a\tau_{\rm
  H}\gg 1$. The dashed lines in  the right panel of Figure
\ref{fig:escape_fraction} show the results for the different
rotational velocities for the homogeneous models. First of all we note
that the escape fraction does not increase significantly from
$\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$. This
counter-intuitive result is a consequence of the transition into the
'extremely' opaque regime,  which occurs roughly when $a\tau_{\rm{H}}
> 10^{3}$  e.g. \citep{Neufeld90}.

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=0.45\textwidth]{nscattvsmu.eps}
    \includegraphics[width=0.45\textwidth]{f4.eps}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  $\mu$ (left panel) and the maximum rotational velocity $V_{\rm
    max}$. The left panel shows the behaviour for $\tau=10^{5}$ and
  $V_{max}=300$\kms but this independence of the independence between
  $N_{\rm scatt}$ and $\mu$ is kept for all models. 
  In the right panel the continuous (dashed) lines represent an
  central (homogeneous) distribution of sources; this shows the clear
  decrease in the number of scatterings for the homogenously
  distributed sources.
\label{fig:Nscatt} }    
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{f5_1.eps}
  \includegraphics[width=0.48\textwidth]{f5_2.eps}
  \includegraphics[width=0.48\textwidth]{f5_3.eps}
  \includegraphics[width=0.48\textwidth]{f5_4.eps}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels correspond to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

%% One possible explanation for the emergence of the single peak in the
%% homogeneous systems is that some photons close to the surface
%% (a sort of photosphere) can escape with a low number of scatterings,
%% allowing them to stay close to the line's center. Increasing the
%% rotational velocity $V_{\rm max}$ reduces the optical depth making the
%% photosphere region effectively larger, increasing the number of
%% photons escaping close to the line's center.

%% However, for the central emission the transition to a single peak is
%% harder to reach in the range of explored parameters. 
%% In this case the average number of scatterings to escape
%% should remain close to constant as the rotational velocity
%% increases. The rise in intensity at the line's center could instead
%% mean that the scattering in a medium with bulk motion are inefficient
%% in driving photons outside that frequency region.

%% In order to explore our interpretation for these two scenarios we now 
%% quantify the effect of rotation on the number of scatterings.

The number of scatterings affects the escape frequency of a \ly
photon. Studying this quantity can help us in claryfying two results
that we have observed for homogeneous models in the previous sections:
the quick emergence of a single peak and the increase in the escape
fraction.

A possible explanation for these two results is that some photons
close to the surface can escape with a low number of scatterings. This
has a two-fold consequence: the escaping photons do not have time to
scatter away from the line's center and also have less chances to find
an absorbing dust grain.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. From the right panel observe that the number of
scatterings and the outgoing angle are independent. This plot
corresponds to the specific case of the central model with $\tau=10^5$ and
$V_{\rm max}=300$\kms; but we have veryfied that this holds for all
models. This allows us to focus to use all the photons in the
simulations regardless of their value of $\cos\theta$.

In the right panel of Figure \ref{fig:Nscatt} we observe how the
average number of scatterings decreases for larger rotational
velocities in the case of homogenous source distributions. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case. For
the central source distribution the average number of scatterings
$\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$.

In static cases, a large value of the optical depth correlates with a
high number of scatterings.  This can be precisely quantified in the
static slab. In this model for centrally emitted sources the average
number of scatterings depends only on the optical depth $\langle
N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}, in the case of homogeneously distributed
sources $\langle N_{\rm   scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}.

We perform a similar quantification of in our speherical setup for the
scatic spheres. We find that for the central model the number of
scatterings is proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.
For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, thisrepresents
factors of $(0.66, 0.59, 0.55)$ lower than the centrally  emitted photons.

For the homogeneous rotating sphere at  $V_{\rm{max}}=300\kms$ we find
that $\langle N_{\rm   scatt}\rangle=(0.38, 0.34, 0.42)\tau_{\rm   H}$
which represents   factors of $(0.26, 0.34, 0.46)$ lower than the
corresponding central case.

In order to gain a deeper understanding of these results  we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we these 
results in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.
